---
title: "Test a Phenomenon"
author: "Scott Brenstuhl"
date: "September 27, 2016"
output: html_document
---


```{r echo=FALSE}
library(tidyverse)

stroop <- read.csv("stroopdata.csv")

```


In this experement we are we are investingating the stroop effect. We will be
testing the effect of congruent and incongruent words on the length of time that
it takes for participants to identify the color that the name of a color is
written in.

##### Q1: What is our independent variable? What is our dependent variable?

Our independent variable is if the color is congruent with the work that is
writen (ex. red would be written in red) or incongruent (ex. the word red being
written in yellow)

Our dependent variable is the amount of time it takes to name the colors in the
lists.

##### Q2: What is an appropriate set of hypotheses for this task? What kind of 
statistical test do you expect to perform? Justify your choices.

Our null hypothese is that a color and the color it is written in matching will
have no affect on the amount of time that it takes participants to state the 
color which it is written in.

Our hypothese is that the color that a color is written in will affect how long
it takes people to state the color it is written in.

I have picked these hypotheses because while it seems most likely that a color 
and the color it is writen in matching would make the task take less time, it
is also possible that it will make it take longer and it will be interesting to
explore both possibilities at once.


##### Q3: Report some descriptive statistics regarding this dataset. Include at
least one measure of central tendency and at least one measure of variability.

```{r echo=FALSE}
M_c <- sum(stroop$Congruent)/length(stroop$Congruent)
M_i <- sum(stroop$Incongruent)/length(stroop$Incongruent)

Med_c <- median(stroop$Congruent)
Med_i <- median(stroop$Incongruent)

```


```{r echo=FALSE}
sd_c <- (stroop$Congruent - M_c) ^2 %>%
            sum() %>%
            sqrt()

sd_i <- (stroop$Incongruent - M_i) ^2 %>%
            sum() %>%
            sqrt()

```



##### Q4: Provide one or two visualizations that show the distribution of the 
sample data. Write one or two sentences noting what you observe about the plot 
or plots.

```{r, echo = FALSE}
tidy_stroop <- gather(stroop)
names(tidy_stroop) <- c("coloring", "time")

ggplot(data = tidy_stroop, aes(x= time)) +
    geom_histogram(binwidth = 2) +
    facet_grid(coloring ~.) +
    labs(title = "Time of Stroop Task by Conguency")
```



```{r, echo = FALSE}
ggplot(data = stroop, aes(x=Congruent, y = Incongruent)) +
    geom_point() +
    geom_smooth(method = "lm")

```


##### Q5: Now, perform the statistical test and report your results. What is 
your confidence level and your critical statistic value? Do you reject the null 
hypothesis or fail to reject it? Come to a conclusion in terms of the experiment
task. Did the results match up with your expectations?



##### Q6 (Optional): What do you think is responsible for the effects observed? 
Can you think of an alternative or similar task that would result in a similar 
effect? Some research about the problem will be helpful for thinking about these
two questions!




