---
title: "Test a Phenomenon"
author: "Scott Brenstuhl"
date: "September 27, 2016"
output: html_document
---


```{r alert=FALSE, message = FALSE, echo = FALSE, warning=FALSE}
library(tidyverse)

stroop <- read.csv("stroopdata.csv")

```


In this experement we are we are investingating the stroop effect. We will be
testing the effect of congruent and incongruent words on the length of time that
it takes for participants to identify the color that the name of a color is
written in.

##### Q1: What is our independent variable? What is our dependent variable?

Our independent variable is if the color is congruent with the word that is
writen (ex. <font color = "red"> red </font>) or incongruent (ex. 
<font color = "blue"> red </font>).

Our dependent variable is the amount of time it takes to name the colors in the
lists.

##### Q2: What is an appropriate set of hypotheses for this task? What kind of statistical test do you expect to perform? Justify your choices.

Our null hypothese is that a color and the color it is written in matching will
have no affect on the amount of time that it takes participants to state the 
color which it is written in.

Our hypothese is that the color that a color is written in will affect how long
it takes people to state the color it is written in.

I have picked these hypotheses because while it seems most likely that a color 
and the color it is writen in matching would make the task take less time, it
is also possible that it will make it take longer and it will be interesting to
explore both possibilities at once.


##### Q3: Report some descriptive statistics regarding this dataset. Include at least one measure of central tendency and at least one measure of variability.

```{r echo=FALSE}
M_c <- (sum(stroop$Congruent)/length(stroop$Congruent)) %>%
        round(2)

M_i <- (sum(stroop$Incongruent)/length(stroop$Incongruent)) %>%
        round(2)

Med_c <- median(stroop$Congruent) %>%
            round(2)

Med_i <- median(stroop$Incongruent) %>%
            round(2)

```

The congruent times have a mean of ```r M_c``` and a median of ```r Med_c```.

While the incongruent times have a mean of ```r M_i``` and a median of 
```r Med_i```. 


```{r echo=FALSE}
sd_c <- (stroop$Congruent - M_c) ^2 %>%
            sum() %>%
            sqrt() %>%
            round(2)

sd_i <- (stroop$Incongruent - M_i) ^2 %>%
            sum() %>%
            sqrt() %>%
            round(2)

```

The conruent times' standard deviation is ```r sd_c```.

The incongruent times' standard deviation is ```r sd_i```.

##### Q4: Provide one or two visualizations that show the distribution of the sample data. Write one or two sentences noting what you observe about the plot or plots.

```{r echo = FALSE}
tidy_stroop <- gather(stroop)
names(tidy_stroop) <- c("coloring", "time")

ggplot(data = tidy_stroop, aes(x= time)) +
    geom_histogram(binwidth = 2) +
    facet_grid(coloring ~.) +
    labs(title = "Time of Stroop Task by Conguency")
```

When we look at the faceted histogram it looks pretty clear that the stroop task
usually takes longer to complete when the coloring is incongruent with the
words. At the far right of the "Incongruent" facet we can also see that the 
incongruency makes the task very difficult for some people.

```{r echo = FALSE}
# ggplot(data = stroop, aes(x=Congruent, y = Incongruent)) +
#     geom_point() +
#     geom_smooth(method = "lm")

```


##### Q5: Now, perform the statistical test and report your results. What is your confidence level and your critical statistic value? Do you reject the null hypothesis or fail to reject it? Come to a conclusion in terms of the experiment task. Did the results match up with your expectations?



```{r echo = FALSE}
n <- nrow(stroop)
d_free <- n-1
D <- (stroop$Congruent - stroop$Incongruent)
M <- mean(D)
s <- (sum((D - M)^2) / d_free) %>%
        sqrt()

# Standard error of mean
SEM <- s/sqrt(n)

# t-statistic
t <- (M_c - M_i) / SEM

# Margin of error at .05
MofE <-2.069 * SEM

CI <- c(M - MofE, M + MofE)

```

Since we are doing a two tailed t-test with alpha = .05 and 23 degrees of 
freedom our t-critical is +/- 2.069.

For the difference in task time between congruent colors and incongruent colors 
we get a t-statistic of ```r t ```. Which is well below the -2.069 t-critical 
value and shows that having congrunent colors lowers the time it takes to 
complete the stroop task a statistically significant amount.

With our confidence intervale we can conclude that on average people will
take somewhere in the range of ```r CI ``` fewer seconds to complete the stroop
task when the colors are congruent.



##### Q6 (Optional): What do you think is responsible for the effects observed? Can you think of an alternative or similar task that would result in a similar effect? Some research about the problem will be helpful for thinking about these two questions!




