---
title: "Test a Phenomenon"
author: "Scott Brenstuhl"
output: html_document
---


```{r alert=FALSE, message = FALSE, echo = FALSE, warning=FALSE}
library(tidyverse)

stroop <- read.csv("stroopdata.csv")

```


In this experiment we are we are investigating the stroop effect. We will be
testing the effect of words being congruent and incongruent with the color they
are written in on the length of time that it takes for participants to identify
the color that the name of a color is written in.

##### Q1: What is our independent variable? What is our dependent variable?

Our independent variable is if the color is congruent with the word that is
written (ex. <font color = "red"> red </font>) or incongruent (ex. 
<font color = "blue"> red </font>).

Our dependent variable is the amount of time it takes to name the colors in the
lists.

##### Q2: What is an appropriate set of hypotheses for this task? What kind of statistical test do you expect to perform? Justify your choices.

Our null hypotheses is that a color and the color it is written in matching will
have no affect on the average population time to state the color which it is 
written in. So the population mean time for congruent words is equal to the
population mean time for incongruent words.

H<sub>0</sub>: μ<sub>congruent</sub> = μ<sub>incongruent</sub>

Our alternate hypothesis is that the color that a color is written in will have
an affect on the average population time of how long it takes people to state 
the color it is written in. So the population mean time for congruent words will
not be equal to the population mean time for incongruent words.

H<sub>1</sub>: μ<sub>congruent</sub> ≠ μ<sub>incongruent</sub>

Since the participants performed the task both with the words being congruent 
and incongruent and since time is a continuous variable, we will use a 
dependent t-test. With there being less than 30 observations we will have it be 
a two tailed test which have been shown to be accurate for small sample sizes.


##### Q3: Report some descriptive statistics regarding this dataset. Include at least one measure of central tendency and at least one measure of variability.

```{r echo=FALSE}
M_c <- (sum(stroop$Congruent)/length(stroop$Congruent)) %>%
        round(2)

M_i <- (sum(stroop$Incongruent)/length(stroop$Incongruent)) %>%
        round(2)

Med_c <- median(stroop$Congruent) %>%
            round(2)

Med_i <- median(stroop$Incongruent) %>%
            round(2)

```

The congruent times have a mean of ```r M_c``` and a median of ```r Med_c```.

While the incongruent times have a mean of ```r M_i``` and a median of 
```r Med_i```. 


```{r echo=FALSE}
var_c <- (stroop$Congruent - M_c) ^2 %>%
            sum()/(length(stroop$Congruent) - 1) 

sd_c <- sqrt(var_c) %>%
            round(2)
# check that 'from scratch' math is right
# sd_c == round(sd(stroop$Congruent),2)

var_i <- (stroop$Incongruent - M_i) ^2 %>%
            sum()/(length(stroop$Incongruent) - 1) 

sd_i <- sqrt(var_i) %>%
            round(2)

# check that 'from scratch' math is right
# sd_i == round(sd(stroop$Incongruent),2)
```

The congruent times' standard deviation is ```r sd_c```.

The incongruent times' standard deviation is ```r sd_i```.

##### Q4: Provide one or two visualizations that show the distribution of the sample data. Write one or two sentences noting what you observe about the plot or plots.

```{r echo = FALSE}
tidy_stroop <- gather(stroop)
names(tidy_stroop) <- c("coloring", "time")

ggplot(data = tidy_stroop, aes(x= time)) +
    geom_histogram(binwidth = 2) +
    facet_grid(coloring ~.) +
    labs(title = "Time of Stroop Task by Conguency")
```

When we look at the faceted histogram it looks pretty clear that the stroop task
usually takes longer to complete when the coloring is incongruent with the
words. At the far right of the "Incongruent" facet we can also see that the 
incongruency makes the task very difficult for some people.

```{r echo = FALSE}
# ggplot(data = stroop, aes(x=Congruent, y = Incongruent)) +
#     geom_point() +
#     geom_smooth(method = "lm")

```


##### Q5: Now, perform the statistical test and report your results. What is your confidence level and your critical statistic value? Do you reject the null hypothesis or fail to reject it? Come to a conclusion in terms of the experiment task. Did the results match up with your expectations?



```{r echo = FALSE}
n <- nrow(stroop)
d_free <- n-1
D <- (stroop$Congruent - stroop$Incongruent)
M <- mean(D)
s <- (sum((D - M)^2) / d_free) %>%
        sqrt()

# Standard error of mean
SEM <- s/sqrt(n)

# t-statistic
t <- (M_c - M_i) / SEM

# Margin of error at .05
MofE <-2.069 * SEM

CI_low <- M - MofE
CI_high <- M + MofE

```

Since we are doing a two tailed t-test with alpha = .05 and ```r d_free``` 
degrees of freedom our t-critical is +/- 2.069.

For the difference in task time between congruent colors and incongruent colors 
we get a t-statistic of ```r round(t, 2) ```. Which is well below the -2.069 
t-critical value and shows that having congruent colors lowers the time it takes
to complete the stroop task a statistically significant amount. So we will 
reject the null hypothesis.

t(```r d_free```) = ```r round(t, 2)```, P < .05 

With our confidence interval we can conclude that on average people will
take somewhere between ```r round(CI_low, 2) ```  and ```r round(CI_high, 2)``` 
fewer seconds to complete the stroop task when the colors are congruent than
when they are incongruent.


##### Q6 (Optional): What do you think is responsible for the effects observed? Can you think of an alternative or similar task that would result in a similar effect? Some research about the problem will be helpful for thinking about these two questions!

It seems that your brain automatically reads the words, even though you are 
trying to name the color so if they are the same there is no conflict and you
can state the color you are thinking when you look at the work. If they are 
incongruent though, you have both the color you read and the color of the text
in your head and you have to sort out which to say which slows you down. 

It would be really interesting to see if this carries over to something like 
having a word super imposed on top of an image and trying to name the image and
if having "car" superimposed on a picture of a truck would cause any difference.



##### Resources
I reviewed these concepts further at the following sites:  
http://support.minitab.com/en-us/minitab/17/topic-library/basic-statistics-and-graphs/hypothesis-tests/basics/null-and-alternative-hypotheses/  
https://statistics.laerd.com/statistical-guides/dependent-t-test-statistical-guide.php  
http://www.measuringu.com/blog/small-n.php

