---
title: "Project 4"
output: html_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/',
                      echo=FALSE, warning=FALSE, message=FALSE)
 
```

```{r libraries}
require(tidyverse)
require(ggmap)
require(stringr)

```

```{r load_data}
accidents <- read_csv("FARS2015NationalCSV/accident.csv")
vehicles <- read_csv("FARS2015NationalCSV/vehicle.csv")
person <- read_csv("FARS2015NationalCSV/person.csv")
```

```{r data prep}
# There can be multiple drunk people but I usually want to know if there are any
accidents$DRUNK <- accidents$DRUNK_DR > 0

# They provide this in a dreadful format burried in the data manual this 
# makes it uasable to map codes to states.
raw_states <- "01 Alabama 30 Montana
02 Alaska 31 Nebraska
03 American Samoa 32 Nevada
04 Arizona 33 New Hampshire
05 Arkansas 34 New Jersey
06 California 35 New Mexico
08 Colorado 36 New York
09 Connecticut 37 North Carolina
10 Delaware 38 North Dakota
11 District of Columbia 39 Ohio
12 Florida 40 Oklahoma
13 Georgia 41 Oregon
14 Guam 42 Pennsylvania
15 Hawaii 43 Puerto Rico
16 Idaho 44 Rhode Island
17 Illinois 45 South Carolina
18 Indiana 46 South Dakota
19 Iowa 47 Tennessee
20 Kansas 48 Texas
21 Kentucky 49 Utah
22 Louisiana 50 Vermont
23 Maine 51 Virginia
24 Maryland 52 Virgin Islands (Since 2004)
25 Massachusetts 53 Washington
26 Michigan 54 West Virginia
27 Minnesota 55 Wisconsin
28 Mississippi 56 Wyoming
29 Missouri"

states <- str_replace(raw_states, ' \\(Since 2004\\)', '') %>%
            str_split('\\s(?=[0-9])') %>%
            unlist() %>%
            str_split_fixed(' ', 2) %>%
            as_data_frame()

names(states) <- c("st_code", "st_name")
states$st_code <- as.numeric(states$st_code)
```

```{r}
head(accidents)
head(vehicles)
```


### How many fatalities occure per accident?

```{r fatalities_per_accident}
ggplot(data = accidents, aes(x = FATALS)) +
    geom_histogram(binwidth = 1) +
    ggtitle("Total fatalities in collision") + 
    scale_x_continuous(breaks = seq(1,10))

```


### How many vehicles are involved in accidents with fatalities?

```{r vehicles_per_accident}

ggplot(data = accidents, aes(x = VE_TOTAL)) +
    geom_histogram(binwidth = 1) +
    ggtitle("Total vehicles in collision")

```

1-3 vehicles involved is so dominat that we cant even see that there are some
rare large pileups ranging all the way to 58 cars involved!

```{r}
table(accidents$VE_TOTAL)
```

### What time of day do accidents occur?
```{r collisions_per_hour}

group_by(accidents, HOUR) %>%
    filter(HOUR != 99) %>%
    summarise(fatalities = sum(FATALS)) %>%
    ggplot(aes(x = HOUR, fatalities)) +
        geom_line()+
        ggtitle("Collision per hour")
```


#### Does it change from month to month?

```{r collisions_per_month}
# fig.height=10
accidents %>%    
    mutate(MONTH = as.character(MONTH)) %>%
    group_by(MONTH, HOUR) %>%
    # 99 means unkown
    filter(HOUR != 99) %>%
    summarise(fatalities = sum(FATALS)) %>%
    ggplot(aes(x = HOUR, fatalities, color = MONTH)) +
        geom_line()+
        #facet_grid(MONTH ~.) +
        ggtitle("Collision per hour by month")

```

#### What about by week?

```{r collisions_per_week}
group_by(accidents, DAY_WEEK, HOUR) %>%
    filter(HOUR != 99) %>%
    summarise(fatalities = sum(FATALS)) %>%
    ggplot(aes(x = HOUR, fatalities)) +
        geom_line()+
        facet_grid(DAY_WEEK ~.) +
        ggtitle("Collision per hour by weekday")

```

### When are drunk driver incidents?

```{r drunk_by_weekday_by_hour}
group_by(accidents, DAY_WEEK, HOUR, DRUNK) %>%
    filter(HOUR != 99) %>%
    summarise(fatalities = sum(FATALS)) %>%
    ggplot(aes(x = HOUR, fatalities)) +
        geom_line()+
        facet_grid(DAY_WEEK ~ DRUNK) +
        ggtitle("Collision per hour by weekeday involving drunk drivers")


```

```{r highest_drunk_by_state}

drunk_by_state <- group_by(accidents, STATE) %>%
                        mutate(in_state = n()) %>%
                        group_by(STATE, DRUNK, in_state) %>%
                        summarise(state_drunk = n()) %>%
                        mutate(percent_drunk = round(
                            (state_drunk/in_state) * 100, 2)) %>%
                        filter(DRUNK == TRUE) %>%
                        ungroup() %>%
                        arrange(percent_drunk)

drunk_by_state <- inner_join(drunk_by_state, states,
                             by = c('STATE' = 'st_code'))

# This could be an good one to really clean up and get the point across
tail(drunk_by_state, 10) %>%
    ggplot(aes(x= reorder(st_name, percent_drunk), y = percent_drunk)) +
        geom_bar(stat = 'identity') +
        coord_flip() +
        labs(title = "Highest percent of fatal accidents involve a drunk driver")

```

```{r lowest_drunk_per_state}
head(drunk_by_state, 10) %>%
    ggplot(aes(x= reorder(st_name, desc(percent_drunk)), y = percent_drunk)) +
        geom_bar(stat = 'identity') +
        coord_flip() +
        labs(title = "Lowest percent of fatal accidents involve a drunk driver")
```


### Where are crashes occuring?

```{r plot_lat_long}
ggplot(data = accidents, aes(x = LONGITUD, y = LATITUDE)) +
           geom_point()
```

Alright the data looks like a smashed map, and if we check the data guide we 
see: 

LONGITUD | Meaning
---------|-------------
DDD.DDDD | Actual Degrees
777.7777 | Not Reported
888.8888 | Not Available (If State Exempt)
999.9999 | Unknown

So we can drop any longitude greater than 0 since valid US locations should be 
negative, then we are ready to plot.

```{r collision_map}
valid_location <- filter(accidents, LONGITUD < 0)

us <- c(left = -125, bottom = 25.75, right = -67, top = 49)
map <- get_stamenmap(us, zoom = 5, maptype = "toner-lite")
ggmap(map)+
    stat_density_2d(data = valid_location, 
                    aes(x=LONGITUD, y= LATITUDE, fill = ..level..), 
                    geom = "polygon", alpha = .3, color = NA) +
    scale_fill_gradient2("Fatalities", 
                         low = "green", mid = "yellow", high = "red", 
                         midpoint = .002)


```

As heatmaps usually seem to turn out this is just a population map. This kind
of validates that traffic fatalities happen where people live... which isn't 
exactly shocking.

```{r drunk_map}
drunk_percent <- mutate(valid_location, lon = round(LONGITUD,1), 
                        lat = round(LATITUDE,1)) %>%
                    group_by(lon, lat, DRUNK_DR) %>%
                    summarise(n = n()) %>%
                    mutate(per_drunk = DRUNK_DR/n)%>%
                    group_by(lon,lat) %>%
                    filter(max(per_drunk) == per_drunk) %>%
                    arrange(lon, lat)

ggmap(map)+
    stat_density_2d(data = drunk_percent, 
                    aes(x=lon, y= lat, fill = ..level..), 
                    geom = "polygon", alpha = .3, color = NA) +
                    scale_fill_gradient2("Fatalities", 
                         low = "green", mid = "yellow", high = "red", 
                         midpoint = .0015)


```

## Person data

```{r person_hist}
person %>%
    filter(AGE < 500) %>%
    ggplot(aes(x=AGE)) +
    geom_histogram()

```



# Ideas to plot

Maps

* Heat map (will just be population map probably) 
* Drunk driving incidents in state / tot in state 
* Drunk driving % by month (does it peak arround holidays?)
* Most dangerious driving days of the year

Person Data:
Age (possible to compair to age distribution of state?)
Compair types of restraints and injury severity
Drinking vs. Age
Drugs vs. age
Drugs by state
Breakdowns of cycalist info
Lag time from crash to death
How offten are they at work
* Is there something interesting here when combined with time of day?
LOCATION for where non-motorists were durrint time of crash

```{r count_plots}
length(list.files("Figs/"))
```





