---
title: "Project 4"
output: html_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/',
                      echo=FALSE, warning=FALSE, message=FALSE)
 
```

```{r libraries}
require(lubridate)
require(stringr)
require(tidyverse)
```

## Traffic Fatality Data

```{r load_data}
person <- read_csv("FARS2015NationalCSV/person.csv")
```

I have decided to explore the 2015 U.S. Department of Transportation Fatality 
Analysis Reporting System (FARS) report. I think this is an important data set
to understand because 35,092 people died on public roadways in 2015 (nearly 100
per day!!) an increase of 7.2% since 2014. DJ Patil and Mark Rosekind of Obama's
administration issued [a public call to action](https://obamawhitehouse.archives.gov/blog/2016/08/29/2015-traffic-fatalities-data-has-just-been-released-call-action-download-and-analyze) asking citizens to help explore the data. 

I was orignally looking into all of the data but this report was quickly 
becoming an unmanagable length so I have re-focused on just the person data 
file. It is my hope that by sheding light on who is involved in these collisions 
the data can become more personal than just "35,092 people died" which is hard
to conceptualize. 

### Understanding the data 

From the FARS person data file, we have information on `r nrow(person)` people 
in `r length(unique(person$ST_CASE))` different colisions that had fatalities
durring 2015.  We have `r length(person)` features about the people involved in
these accidents. The in depth details of each field is available in the [FARS
Analytical User Guide](ftp://ftp.nhtsa.dot.gov/fars/FARS-DOC/Analytical%20User%20Guide/USERGUIDE-2015.pdf)
on pages 240 - 284, I will breifly touch on all the fields as we explore them 
though so don't dive into read the documentation yet!

To get a general understanding, here are all the fields:

```{r base_features}
names(person)
```

As you may be able to tell, these can be broken down into a few different types
of catagories info about the crash in general (STATE, DAY, HOUR, ect), info 
about the vehicle they were in (MAKE, BODY_TYP, etc.), info about the 
individual's situation in the collision (SEAT_POS, INJ_SEV, EJECTION, DRUGS, 
etc.), and demographics (AGE, SEX, RACE, etc.).


```{r new_features}
# In the userguide, death dates is the most clearly labled indicator of
# non-fatal involvment
person$SURVIVED <- person$DEATH_MO == 88

# More converniant date format
person$COLLISION_DATE <- ymd(paste0(2015, "/", person$MONTH, "/", person$DAY))

# More converniant datetime format if hour or min is unknown this is NA
person$COLLISION_DATETIME <- ymd_hm(paste0(2015, "/", 
                                       person$MONTH, "/",
                                       person$DAY, "/",
                                       person$HOUR, "/",
                                       person$MINUTE))

# They provide this in a dreadful format burried in the data manual this 
# makes it uasable to map codes to states.
raw_states <- "01 Alabama 30 Montana
02 Alaska 31 Nebraska
03 American Samoa 32 Nevada
04 Arizona 33 New Hampshire
05 Arkansas 34 New Jersey
06 California 35 New Mexico
08 Colorado 36 New York
09 Connecticut 37 North Carolina
10 Delaware 38 North Dakota
11 District of Columbia 39 Ohio
12 Florida 40 Oklahoma
13 Georgia 41 Oregon
14 Guam 42 Pennsylvania
15 Hawaii 43 Puerto Rico
16 Idaho 44 Rhode Island
17 Illinois 45 South Carolina
18 Indiana 46 South Dakota
19 Iowa 47 Tennessee
20 Kansas 48 Texas
21 Kentucky 49 Utah
22 Louisiana 50 Vermont
23 Maine 51 Virginia
24 Maryland 52 Virgin Islands (Since 2004)
25 Massachusetts 53 Washington
26 Michigan 54 West Virginia
27 Minnesota 55 Wisconsin
28 Mississippi 56 Wyoming
29 Missouri"

states <- str_replace(raw_states, ' \\(Since 2004\\)', '') %>%
            str_split('\\s(?=[0-9])') %>%
            unlist() %>%
            str_split_fixed(' ', 2) %>%
            as_data_frame()

names(states) <- c("st_code", "STATE_NAME")
states$st_code <- as.numeric(states$st_code)

person <- left_join(person, states, by = c("STATE" = "st_code"))

```



To get started I have created four new features. SURVIVED is a clear 
lableing of if the person survived or died in the collision. With our data set
having `r sum(!person$SURVIVED)` fatalities which matches the 35,092 released 
by the whitehouse, we can validate our data somewhat. I also added
COLLISION_DATE and COLLISION_DATETIME fields in more usable date formats in
case they will be useful later. Finally I created the STATE_NAME column by 
mapping the state ids that are in the dataset to the coresponding state from 
the documentation.

There is obviously tons to be explored, so let's dive in!

### Who's involved in these colisions?

```{r people_per_crash}

person %>%
    group_by(ST_CASE) %>%
    summarise(people_involved = n()) %>%
    ggplot(aes(x = people_involved)) +
        geom_histogram(binwidth = 1)

```

It appears that it is most common for there to be two poeople involved in a 
collision whether is be two people in one vehicle, one in each of two vehicles,
or one in a vehicle and a pedestrian. As you might expect the more people 
involved in one collsion, the less rare it is althought there seems to have been
at least one collison with almost 100 people involved.

```{r veh_per_crash}

veh_per_crash <- person %>%
                    group_by(ST_CASE) %>%
                    summarise(vehicle_per_collision = max(VEH_NO)) 
veh_per_crash %>%
    ggplot(aes(x = vehicle_per_collision)) +
        geom_histogram(binwidth = 1)
```

It seems that single vehicle crashes are most common.

```{r no_vehicles}
no_vehicle <- filter(veh_per_crash, vehicle_per_collision == 0)[['ST_CASE']]

#View(filter(person, ST_CASE %in% no_vehicle))

```
 
I found it odd that there were any collisons with 0 vehicles but after looking
at just casses that had no vehicles (Icommented out the View statement but 
feel free to look yourself)  it looks like they are all missing
all vehicle information and the PER_TYPE labels them as non-motor vehicle 
occupants. My working suspicion is that these are pedestrians that were struck
in a hit and run so the individual was found but the vehicle that struck them
was not.

```{r ppl_per_veh}
person %>%
    group_by(ST_CASE,VEH_NO) %>%
    summarise(person_per_vehicle = max(PER_NO)) %>%
    ggplot(aes(x = person_per_vehicle)) +
        geom_histogram(binwidth = 1)
```

Finally, it is most common for there to be one person in each vehicle.

```{r age_hist}
                

age_hist <- function(people){
                people %>%
                filter(AGE < 500) %>%
                ggplot(aes(x=AGE, fill = !SURVIVED)) +
                geom_histogram(binwidth = 5) +
                scale_x_continuous(breaks = seq(0,120,10))
            }

person %>%
    age_hist

```

I'm starting to see why my insurance was so high in my teens and early twenties,
there are many more people that age involved in collisions than older adults.
I have also colored the histrogram with the inverse of SURVIVED so the we can
easily see the distribution of people that passed away in these accidents.
However this would include passengers, let's focus on drivers.

```{r drivers_age_hist}
# Saving since will likely use this often 
drivers <- filter(person, SEAT_POS == 11)

drivers %>%
    age_hist()

```

This actually flattens the distribution out a good bit which leads me to 
believe that younger people thend to ride along with others more often than
older people do.

Speaking of insurance... 

```{r driver_age_sex_hist}

sex_label <- tibble(sex_id = c(1,2),
                    sex_name = c('Male', 'Female'))
drivers %>%
    # There are not enough 8 or 9s to be visable
    filter(SEX < 3) %>%
    left_join(sex_label, by = c("SEX" = "sex_id")) %>%
    age_hist() +
    facet_grid(sex_name~ .)
```

I can also see why being male leads to higher premiums! It is shocking 
how much more often males are drivers in fatal accidents. It looks to be at 
least twice as common for most ages.

### When do accidents occure?
```{r most_dangerious_days}
# There is one record per person, this gives one per crash instead
individual_crashes <- function(people){
    people %>%
        select(ST_CASE, MONTH, DAY, HOUR, MINUTE, COLLISION_DATE) %>%
        distinct()
}

person %>%
    individual_crashes() %>%
    group_by(COLLISION_DATE) %>%
    summarise(crash_per_day = n()) %>%
    arrange(desc(crash_per_day)) %>%
    head(10) %>%
    mutate(weekday = wday(COLLISION_DATE, TRUE))
```

Besides all being weekend days, there doesnt seem to be anything special about
the ten days with the most collisions.

```{r avg_crash_per_weekday}
person %>%
    individual_crashes() %>%
    mutate(weekday = wday(COLLISION_DATE, TRUE)) %>%
    group_by(COLLISION_DATE, weekday) %>%
    summarise(daily_crashes = n()) %>%
    group_by(weekday) %>%
    summarise(avg_weekday_crashes = sum(daily_crashes)/n()) %>%
    ggplot(aes(x = weekday, y = avg_weekday_crashes)) +
        geom_bar(stat='identity')
        
```

Along the same lines, Saturadays and Sundays seem to have the most collisions
followed by Fridays.

```{r crashes_by_month}
crashes_over_time <- function(people, time_interval, facet_col = NA){    
                            people %>%
                            # this needs to be pasted together to work with
                            # lazy evaluation correctly
                            filter_(paste0(time_interval , "!=", 99)) %>%
                            group_by_(time_interval, facet_col) %>%
                            summarise(crashes = n()) %>%
                            ggplot(aes_string(x = time_interval, 
                                              y = "crashes")) +
                                geom_line()
                        }
        

person %>%
    individual_crashes %>%
    crashes_over_time('MONTH') +
    scale_x_continuous(breaks = seq(0,12,2))
        
        
```

I was curious to see if there are more collisions in November/December than 
other months since there are so many people traveling plus celibration drinks,
this isnt so strightforward though since different months have different number 
of days.

```{r avg_crashes_per_day_by_month}
person %>%
    individual_crashes %>%
    group_by(MONTH) %>%
    summarise(crashes = n()) %>%
    mutate(crash_per_day = crashes/days_in_month(MONTH)) %>%
    ggplot(aes(as.factor(MONTH), crash_per_day)) +
        geom_bar(stat = 'identity') 

```

Interestingly, while this does flatten out the plot it does't seem like the 
holiday months are especially high as I had expected. It almost looks more like
in the colder months of the year people might not be out driving as much?

```{r crashes_by_hour}
person %>%
    individual_crashes %>%
    crashes_over_time('HOUR') +
        scale_x_continuous(breaks = seq(0, 24, 4)) +
        ggtitle("Collision per hour")

```


In the early morning hours crashes are at their lowest then they raise while 
people are commuting to work before dropping back down and then stedily rising 
throughout the day with the most dangerious time being 6pm while people are 
driving home from work. 


```{r crashes_by_hour_weekday, fig.height= 12}
person %>%
    individual_crashes %>%
    mutate(weekday = wday(COLLISION_DATE, TRUE)) %>%
    crashes_over_time('HOUR', 'weekday') +
    facet_grid(weekday ~ .)
```

By looking at the hours crashes occur by hour by day, it is very obvious that 
there are a lot of crashes on late on Fridays and Saturdays and early on 
Saturdays and Sundays.

```{r working_crashes_by_hour_weekday, fig.height= 12}
person %>%
    filter(WORK_INJ == 1) %>%
    individual_crashes %>%
    mutate(weekday = wday(COLLISION_DATE, TRUE)) %>%
    crashes_over_time('HOUR', 'weekday') +
    facet_grid(weekday ~ .)
```

WORK_INJ only shows when the person's death certificate says that they were 
"at work" at the time of the crash. As you may expect, there seem to be the most
incedents where people are working durring the day on weekdays.

### Drug and Alcohol Use 

```{r}

tibble(
    persons_state = c('drunk_driver', 'sober_driver', 'drunk_passenger', 'sober_passenger'),
    count = with(person, c(sum(DRINKING == 1 & SEAT_POS == 11),
            sum(DRINKING == 0 & SEAT_POS == 11),
            sum(DRINKING == 1 & SEAT_POS != 11),
            sum(DRINKING == 0 & SEAT_POS != 11)))
) %>%
    ggplot(aes(x = persons_state, y = count)) +
    geom_bar(stat = 'identity')
    
```

I found it odd that there were so many drunk and sober passengers
but it has occured to me that perhaps police officers on the scene of a fatal 
accident have better things to do than check if passengers have been drinking.

```{r perc_alc_reported}
perc_alc_reported <- filter(person, SEAT_POS != 11) %>%
                        summarize(perc_alc_reported = sum((DRINKING <= 1)) / n()) %>%
                        round(4) * 100

```

This seems somewhat confirmed by only `r perc_alc_reported`% of passengers 
having either alcohol involved or alcohol not involved reported for them. 

```{r drunk_v_not_by_hour}
person %>%
    filter(SEAT_POS == 11, DRINKING <= 1, HOUR < 25) %>%
    group_by(HOUR, DRINKING) %>%
    summarise(count = n()) %>%
    ggplot(aes(x = HOUR, y = count, color = factor(DRINKING))) +
    geom_line()

```

I suppose it makes sense for there to be less drunk driving incidents in the 
middle of the day than in the evenings, but I am a bit surprised by how 
strongly accidents where driver drinking was a factor ran inversely to those
where it wasn't.

```{r drunk_v_not_by_hour_by_day}
person %>%
    filter(SEAT_POS == 11, DRINKING <= 1, HOUR < 25) %>%
    mutate(weekday = wday(COLLISION_DATE, TRUE)) %>%
    group_by(HOUR, weekday, DRINKING) %>%
    summarise(count = n()) %>%
    ggplot(aes(x = HOUR, y = count, color = factor(DRINKING))) +
    geom_line() +
    facet_grid(weekday~.)


```

By looking at it per day, you get basically what you would expect with there 
being a slight increase in drunk driver crashes areound dinner and happy hour 
time throughout the week, but with the big spikes in drunk driving being on the
weekend late night/early mornings. 

Another thing I hadnt caught in the earlier plots of time per weekday is that
you can basically see people sleeping in on the weekend morning compaired to 
weekdays.

```{r}

person %>%
    filter(DRINKING == 1 | DRUGS == 1, HOUR < 25) %>%
    mutate(impairment = if_else(DRUGS == 1 & DRINKING == 1,
                                "Drugs and Dinking", 
                                if_else(DRUGS == 1, 'Drugs', 'Drinking')),
           weekday = wday(COLLISION_DATE, TRUE)) %>%
    group_by(HOUR, weekday, impairment) %>%
    summarise(count = n()) %>%
    ggplot(aes(x = HOUR, y = count, color = factor(impairment))) +
    geom_line() +
    facet_grid(weekday~.)

```

This is a really interesting one to unpack. I am pretty surprised to see how 
often there are more drugs and drinking than people only using drugs. It's also
interesting to me that while durring the day all substance abuse collisions are
a lot lower, almost no one is doing drungs and drinking mid-weekday (althought 
there are still accidents with one or the other).

### Child Restraints

```{r child_restraints}
child_restraints <- tibble(id = c(4,10,11,12), 
                           type = c("Child Restraint Type Unknown",
                                    "Forward Facing",
                                    "Rear Facing",
                                    "Booster Seat"))

child_restraint_used <- person %>%
                            filter(REST_USE %in% child_restraints$id) %>%
                            group_by(REST_USE) %>%
                            summarise(perc_survived = sum(SURVIVED)/n())

child_restraint_used <- inner_join(child_restraint_used, child_restraints, 
                                    by = c("REST_USE" = "id"))

ggplot(child_restraint_used, aes(x = type, y= perc_survived)) +
    geom_bar(stat = 'identity')


```
It's a relief to see that children seats do seem to keep children somewhat 
safe, compaired the the `r round(sum(person$SURVIVED)/nrow(person) * 100, 2)`%
survival rate for everyone that is involved in crashes that have a fatality.
Althought with the amount of collisions that only have one person in each 
vehicle this comparison can't really be made so directly.


```{r child_seat_pos}
seat_pos <- tibble(id = c(0, 11, 12, 13, 18, 19, 21, 22, 23, 28, 29, 31, 32,
                          33, 38, 39, 41, 42, 43, 48, 49, 50, 51, 98, 99),
                   pos = c("Not a Motor Vehicle Occupant",
                            "Front Seat – Left Side",
                            "Front Seat – Middle",
                            "Front Seat – Right Side",
                            "Front Seat – Other",
                            "Front Seat – Unknown",
                            "Second Seat – Left Side",
                            "Second Seat – Middle",
                            "Second Seat – Right Side",
                            "Second Seat – Other",
                            "Second Seat – Unknown",
                            "Third Seat – Left Side",
                            "Third Seat – Middle",
                            "Third Seat – Right Side",
                            "Third Seat – Other",
                            "Third Seat – Unknown",
                            "Fourth Seat – Left Side",
                            "Fourth Seat – Middle",
                            "Fourth Seat – Right Side",
                            "Fourth Seat – Other",
                            "Fourth Seat – Unknown",
                            "Sleeper Section of Cab",
                            "Other",
                            "Not Reported",
                            "Unknown")
                    )

child_restraint_used_pos <- person %>%
                            filter(REST_USE %in% child_restraints$id) 

# Rows are not reported directly so we need to join with seat positions and use regex to get row
child_restraint_used_pos <- left_join(child_restraint_used_pos, seat_pos,
                                      by = c("SEAT_POS" ="id")) %>%
                                left_join(child_restraints, by = c("REST_USE" = "id"))

child_restraint_used_pos$SEAT_ROW <- gsub("(Seat).*", "", child_restraint_used_pos$pos)

ggplot(child_restraint_used_pos, aes(x =SEAT_ROW, fill = type)) +
    geom_bar() +
    labs(title= "Child Seat Row by Restraint Type") +
    coord_flip()

```

On the bright side, it appers that most peopel are avoiding putting child seats
in the front rows of vehicles!

```{r under_10_retraints}
all_restraints <- tibble(
        id = c(00, 02, 01, 03, 04, 05, 07, 08, 10, 11, 12, 16, 17, 19),
        restraint = c('Not Applicable',
                    'Shoulder Belt Only Used',
                    'Lap Belt Only Used',
                    'Lap and Shoulder Belt Used',
                    'Child Restraint Type Unknown ',
                    'DOT-Compliant Motorcycle Helmet ',
                    'None Used',
                    'Restraint Used – Type Unknown',
                    'Child Restraint System – Forward Facing',
                    'Child Restraint System – Rear Facing ',
                    'Booster Seat',
                    'Helmet, Other than DOT-Compliant Motorcycle Helmet',
                    'No Helmet',
                    'Helmet, Unknown if DOT-Compliant')
)

under_10 <- person %>%
    # REST_USE 96 is "Not a Motor Vehicle Occupant"
    filter(AGE <= 10, !(REST_USE %in% c(00,96,98,99,97))) %>%
    mutate(child_restraint_used = REST_USE %in% child_restraints$id) %>%
    left_join(all_restraints,  by = c("REST_USE" = "id"))

under_10 %>%
    group_by(restraint) %>%
    summarise(count = n()) %>%
    ggplot(aes(x = restraint, y = count)) + 
    geom_bar(stat = 'identity') +
    coord_flip()
    
```

I'm honestly shocked by how many kids under 10 are in cars with just a seatbelt
or even more terrifyingly with no restraint at all.

```{r under_10_no_restraint_by_age}

under_10 %>%
    mutate(not_ch_rest = !(REST_USE %in% child_restraints$id)) %>%
    group_by(AGE) %>%
    summarise(perc_by_age = sum(not_ch_rest)/n()) %>%
    ggplot(aes(x = AGE, y = perc_by_age)) +
    geom_line() +
    scale_y_continuous(breaks = seq(0,1,.1))
    
```

While most of my Googling has suggested that children should be in a booster
seat until they are 8-10 years old it looks like it's actually rare for
people to use child restraints that long.

```{r survival_rates_under_10_restraints}
under_10 %>%
    mutate(ch_res_used = REST_USE %in% child_restraints$id) %>%
    group_by(AGE, ch_res_used, SURVIVED) %>%
    summarize(count = n()) %>%
    mutate(perc_survived = count/sum(count)) %>%
    filter(SURVIVED == TRUE) %>%
    ggplot(aes(x=AGE, y = perc_survived, color = ch_res_used)) +
    geom_point()

```

With a higher percent of children surviving at every age if they were in a 
child restraint, it seems like we cant stress enough how important they are!

In an atempt not to sensationalize the 100% of 10 year olds in child restraints 
surviving, it should be noted that there were only
`r nrow(filter(under_10, REST_USE %in% child_restraints$id, AGE == 10))` of
them while there were
`r nrow(filter(under_10, !(REST_USE %in% child_restraints$id), AGE == 10))`
not in restraints. 

### Lag time

```{r lag_time}
person %>%
    filter(LAG_HRS < 900, AGE < 900, DOA == FALSE) %>%
    ggplot(aes(AGE, (LAG_HRS*60) + LAG_MINS)) +
        geom_point(alpha = .2) +
        geom_smooth(method='lm')
    
```

I originally thought that lag time was how long it took for an individual to 
get to the hospital. Unfortunatly this field is a bit darker, since it actually
is how long the person survived from the crash until they eventually passed 
away.

I thought it was interesting that the older one is, the longer they may 
survive before passing away. However there are tons of confounding factors with 
this for instance someone who is younger may be in a similar physical situation
to someone older but the younger person may recover, which would mean they do
not have a lag time. So I don't think we can draw anything too interesting 
from this.

```{r}
names(person)
```

## Final Plots and Summary

## Reflection

# Ideas to plot

Maps

* Heat map (will just be population map probably) 
* Drunk driving incidents in state / tot in state 
* Drunk driving % by month (does it peak arround holidays?)
* Most dangerious driving days of the year

Person Data:
Age (possible to compair to age distribution of state?)
Compair types of restraints and injury severity
Drinking vs. Age
Drugs vs. age
Underage drunks
Drugs by state
Breakdowns of cycalist info
Lag time from crash to death
How offten are they at work
* Is there something interesting here when combined with time of day?
LOCATION for where non-motorists were durrint time of crash
Survival by seat
Number of people per car
number of people per accident
AVG number of people in car depending on age

Can I plot position in the car by making factors for row and left, middle, right and
adding jitter and alpha to understand where people sit?

color for survived vs no plot lag time

```{r count_plots}
length(list.files("Figs/"))
```
